---
title: "NBA Salaries exploration"
author: "John Yuill"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE,
                      message=FALSE,
                      warning=FALSE)

## libraries
library(tidyverse)
library(scales)
library(plotly)

## viz settings
th1 <- theme_classic()

```

## Exploring descriptive stats on NBA salaries

```{r IMPORT DATA}
nba_sal <- read_csv("input/nba-salary.csv")

```

### Overall NBA

Summarize key fields

```{r}
nba_sum <- nba_sal %>% select(salary, season_end, season_start)
summary(nba_sum)
```

What are overall trends in average salary?

```{r}
nba_avg_sal <- nba_sum %>% group_by(season_start) %>% 
  summarize(avg_sal=mean(salary),
            med_sal=median(salary))
```

```{r}
## BASIC plot with two lines: one for avg, one for med
## specify colors within aes to get legend
# ggplot(nba_avg_sal, aes(x=season_start, y=avg_sal))+geom_line(aes(color='avg_sal'))+
#   geom_line(aes(y=med_sal, color='med_sal'))+
#   scale_y_continuous(label=comma)

## ALTERNATIVE - gather metrics
## advantage: simpler code for plot, cleaner legend
nba_sal_gather <- nba_avg_sal %>% gather(key="metric", val="sal", -season_start)

ggplot(nba_sal_gather, aes(x=season_start, y=sal, color=metric))+
  geom_line()+
  scale_y_continuous(label=comma)+
  theme_classic()

```

What does the distribution look like over the years?

Boxplot view

```{r}
nba_sum$season_start <- as.factor(nba_sum$season_start) 
ggplot(nba_sum, aes(x=season_start, y=salary))+geom_boxplot()+
  scale_y_continuous(labels=comma)+
  th1

```

Histogram view

```{r, fig.height=12}

#ggplot(nba_sum, aes(x=salary))+geom_histogram()+
#  facet_wrap(.~season_start)

ggplot(nba_sum, aes(x=salary))+geom_histogram()+
  facet_grid(season_start~.)+
  scale_x_continuous(labels=comma)

```

Too many years to deal with. How about group into 5 yrs.

```{r}
## use starting yr of 5 yr period
nba_sum5 <- nba_avg_sal
nba_sum5$fiveyr <- NA

## set start row s plus 4 rows after f to get 5 rows
## set fiveyr var for lowest season_start yr in set
## increase s to s +5 to jump to next 5 rows
## repeat until end

s <- 1
f <- s+4
while(s<nrow(nba_sum5)){
  nba_sum5$fiveyr[c(s:f)] <- nba_sum5$season_start[s]
s <- s+5
f <- min(s+4,nrow(nba_sum5))
}
nba_sum5 <- nba_sum5 %>% select(season_start, fiveyr)
nba_sum5$season_start <- as.factor(nba_sum5$season_start)
## join

nba_sum5yr <- left_join(nba_sum, nba_sum5, by='season_start')

```

Better view when grouped over 5 yrs. 

```{r, fig.height=10}
ggplot(nba_sum5yr, aes(x=salary))+geom_histogram()+
  facet_grid(fiveyr~.)+
  scale_x_continuous(labels=comma)
```

### Scottie Pippin

Scottie Pippin is considered to have established himself as one of the all-time greats while at Chicago Bulls in the 1990's. At the same time, reported to have relatively low salary.

#### Career

The data I' using has player names as 'player_id'. Luckily, his is easy to find.

```{r}
nba_pip <- nba_sal %>% filter(str_detect(player_id,'pipp'))
nba_pip[,c(2:4,7)]
```

```{r}
ggplot(nba_pip, aes(x=season_start, y=salary, fill=team))+geom_col()+
  scale_y_continuous(labels=comma)+th1

```

Looks like even in early years with the Bulls, his salary varied.

Where does he rank in the league each year?

```{r RANK}
## rank players by salary for each yr
nba_sal_rank <- nba_sal %>% group_by(season_start) %>% mutate(
  sal_rank=rank(desc(salary))
)
## test rank for a given yr
nba_sal_rank_yr <- nba_sal_rank %>% filter(season_start==1990) %>%
  arrange(sal_rank)
```

```{r PIPPEN RANK}
nba_pip_rank <- nba_sal_rank %>% filter(str_detect(player_id,'pipp'))

ggplot(nba_pip_rank, aes(x=season_start, y=sal_rank, fill=team))+geom_col()+
  th1
```

Varies wildly from year to year, with early years getting progressively bad, the rectified for 1991-1993, but started slipping again, until late 1990s. Houston Rockets and especially Portland Trail Blazers were much better to Scottie than the Bulls!

Where does he rank percentile-wise?

```{r}
## filter for years Pippen played
nba_sal_pipyr <- nba_sal %>% filter(season_start>=1987 & season_start<=2004)

## convert season_start fields to factor for boxplot
nba_sal_pipyr$season_start <- as.factor(nba_sal_pipyr$season_start)
nba_pip$season_start <- as.factor(nba_pip$season_start)

ggplot(nba_sal_pipyr, aes(x=season_start, y=salary))+geom_boxplot()+
  #geom_hline(yintercept=nba_pip$salary)+ ## draws lines across chart
  geom_point(data=nba_pip, aes(x=season_start, y=salary), 
             shape=19, color='red', size=2)+
  scale_y_continuous(labels=comma)+
  th1
```

