---
title: "NBA Salaries exploration"
author: "John Yuill"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE,
                      message=FALSE,
                      warning=FALSE)

## libraries
library(tidyverse)
library(scales)
library(plotly)

## viz settings
th1 <- theme_classic()

```

## Exploring descriptive stats on NBA salaries

```{r IMPORT DATA}
nba_sal <- read_csv("input/nba-salary.csv")

```

### Overall NBA

Summarize key fields

```{r}
nba_sum <- nba_sal %>% select(salary, season_end, season_start)
summary(nba_sum)
```

#### What are overall trends in average salary?

```{r AVE SAL}
## group by season to get annual average and median
## add in sd and mean +/- 1 sd for further investigation
nba_avg_sal <- nba_sum %>% group_by(season_start) %>% 
  summarize(avg_sal=mean(salary),
            med_sal=median(salary),
            sd_sal=sd(salary),
            avg_sd_one_low=avg_sal-sd_sal,
            avg_sd_one_high=avg_sal+sd_sal)
```

```{r}
## BASIC plot with two lines: one for avg, one for med
## specify colors within aes to get legend
# ggplot(nba_avg_sal, aes(x=season_start, y=avg_sal))+geom_line(aes(color='avg_sal'))+
#   geom_line(aes(y=med_sal, color='med_sal'))+
#   scale_y_continuous(label=comma)

## ALTERNATIVE - gather metrics
## advantage: simpler code for plot, cleaner legend
nba_sal_gather <- nba_avg_sal %>% gather(key="metric", val="sal", -season_start)

ggplot(nba_sal_gather, aes(x=season_start, y=sal, color=metric))+
  geom_line()+
  scale_y_continuous(label=comma)+
  theme_classic()

```

#### What does the distribution look like over the years?

Boxplot view

```{r BOXPLOT}

## boxplot
## need to set x axis as factor for boxplot
nba_sum$season_start <- as.factor(nba_sum$season_start) 

ggplot(nba_sum, aes(x=season_start, y=salary))+geom_boxplot()+
  scale_y_continuous(labels=comma)+
  th1
```

```{r BOXPLOT SD }
## set x axis from reference data to factor 
nba_avg_sal$season_start <- as.factor(nba_avg_sal$season_start) 

ggplot(nba_sum, aes(x=season_start, y=salary))+geom_boxplot()+
  geom_point(data=nba_avg_sal, aes(x=season_start, y=avg_sal), 
             shape=19, color='blue', size=2)+
  geom_point(data=nba_avg_sal, aes(x=season_start, y=avg_sd_one_low), 
             shape=19, color='red', size=2)+
  geom_point(data=nba_avg_sal, aes(x=season_start, y=avg_sd_one_high), 
             shape=19, color='green', size=2)+
  scale_y_continuous(labels=comma)+
  th1
```


Histogram view

```{r, fig.height=12}

#ggplot(nba_sum, aes(x=salary))+geom_histogram()+
#  facet_wrap(.~season_start)

ggplot(nba_sum, aes(x=salary))+geom_histogram()+
  facet_grid(season_start~.)+
  scale_x_continuous(labels=comma)

```

Too many years to deal with. How about group into 5 yrs.

```{r}
## use starting yr of 5 yr period
nba_sum5 <- nba_avg_sal
nba_sum5$fiveyr <- NA

## set start row s plus 4 rows after f to get 5 rows
## set fiveyr var for lowest season_start yr in set
## increase s to s +5 to jump to next 5 rows
## repeat until end

s <- 1
f <- s+4
while(s<nrow(nba_sum5)){
  nba_sum5$fiveyr[c(s:f)] <- nba_sum5$season_start[s]
s <- s+5
f <- min(s+4,nrow(nba_sum5))
}
nba_sum5 <- nba_sum5 %>% select(season_start, fiveyr)
nba_sum5$season_start <- as.factor(nba_sum5$season_start)
## join

nba_sum5yr <- left_join(nba_sum, nba_sum5, by='season_start')

```

Better view when grouped over 5 yrs. 

```{r, fig.height=10}
ggplot(nba_sum5yr, aes(x=salary))+geom_histogram()+
  facet_grid(fiveyr~.)+
  scale_x_continuous(labels=comma)
```

As shown with boxplot, the center shifts up a little over the years - maybe not that much different from inflation - but the tail gets stretched longer over time.

### Scottie Pippin

Scottie Pippin is considered to have established himself as one of the all-time greats while at Chicago Bulls in the 1990's. At the same time, reported to have relatively low salary.

#### Career

The data I' using has player names as 'player_id'. Luckily, his is easy to find.

```{r}
nba_pip <- nba_sal %>% filter(str_detect(player_id,'pipp'))
nba_pip[,c(2:4,7)]
```

```{r}
ggplot(nba_pip, aes(x=season_start, y=salary, fill=team))+geom_col()+
  geom_hline(yintercept = 1000000, linetype='dashed')+
  scale_y_continuous(labels=comma, expand=c(0,0))+th1

```

(dotted line = $1M)

Looks like even in early years with the Bulls, his salary varied. Although pushing $1M and then jumping well above it.

#### Where does he rank in the league each year?

```{r RANK}
## rank players by salary for each yr
nba_sal_rank <- nba_sal %>% group_by(season_start) %>% mutate(
  sal_rank=rank(desc(salary))
)
## test rank for a given yr
nba_sal_rank_yr <- nba_sal_rank %>% filter(season_start==1990) %>%
  arrange(sal_rank)
```

```{r PIPPEN RANK}
nba_pip_rank <- nba_sal_rank %>% filter(str_detect(player_id,'pipp'))

ggplot(nba_pip_rank, aes(x=season_start, y=sal_rank, fill=team))+geom_col()+
  th1
```

Varies wildly from year to year, with early years getting progressively bad, then rectified for 1991-1993, but started slipping again, until late 1990s. Houston Rockets and especially Portland Trail Blazers were much better to Scottie than the Bulls!

How does he compare to Jordan, say in 1990?

```{r}
nba_rank_compareyr <- nba_sal_rank %>% filter(season_start==1990)
nba_rank_compare <- nba_rank_compareyr %>% filter(sal_rank==1 | str_detect(player_id,'pipp') | str_detect(player_id,'jord'))

nba_rank_compare[,c(2:4,7,8)]
```

```{r}
ggplot(nba_rank_compare, aes(x=player_id, y=salary))+geom_col()+
  geom_hline(yintercept=median(nba_rank_compareyr$salary), linetype='dashed')+
  scale_y_continuous(labels=comma, expand=c(0,0))+
th1
```

(dotted line = 1990 average salary)

Pippen slightly above average but well below his comrade Jordan.

#### Where does he rank percentile-wise?

```{r}
## filter for years Pippen played
nba_sal_pipyr <- nba_sal %>% filter(season_start>=1987 & season_start<=2004)

## convert season_start fields to factor for boxplot
nba_sal_pipyr$season_start <- as.factor(nba_sal_pipyr$season_start)
nba_pip$season_start <- as.factor(nba_pip$season_start)

ggplot(nba_sal_pipyr, aes(x=season_start, y=salary))+geom_boxplot()+
  #geom_hline(yintercept=nba_pip$salary)+ ## draws lines across chart
  geom_point(data=nba_pip, aes(x=season_start, y=salary), 
             shape=19, color='red', size=2)+
  scale_y_continuous(labels=comma)+
  th1
```

How does he compare to Jordan?

```{r}

```

