---
title: "NBA Salaries exploration"
author: "John Yuill"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE,
                      message=FALSE,
                      warning=FALSE)

## libraries
library(tidyverse)

library(scales)
library(plotly)

## viz settings
th1 <- theme_classic()
## OR just set theme
theme_set(theme_classic())

```

## Exploring descriptive stats on NBA salaries

```{r IMPORT DATA}
nba_sal <- read_csv("input/nba-salary.csv")

```

### Overall NBA

**Summarize key fields**

```{r}
nba_sum <- nba_sal %>% select(salary, season_end, season_start)
summary(nba_sum)
```

#### What are overall trends in average salary?

```{r AVE SAL}
## group by season to get annual average and median
## add in sd and mean +/- 1 sd for further investigation
nba_avg_sal <- nba_sum %>% group_by(season_start) %>% 
  summarize(avg_sal=mean(salary),
            med_sal=median(salary),
            sd_sal=sd(salary),
            avg_sd_one_low=avg_sal-sd_sal,
            avg_sd_one_high=avg_sal+sd_sal)
```

```{r}
## BASIC plot with two lines: one for avg, one for med
## specify colors within aes to get legend
# ggplot(nba_avg_sal, aes(x=season_start, y=avg_sal))+geom_line(aes(color='avg_sal'))+
#   geom_line(aes(y=med_sal, color='med_sal'))+
#   scale_y_continuous(label=comma)

## ALTERNATIVE - gather metrics
## advantage: simpler code for plot, cleaner legend
nba_sal_gather <- nba_avg_sal %>% gather(key="metric", val="sal", -season_start)
nba_sal_gather <- nba_sal_gather %>% filter(metric=='avg_sal' | metric=='med_sal')

ggplot(nba_sal_gather, aes(x=season_start, y=sal, color=metric))+
  geom_line()+
  scale_y_continuous(label=comma)

```

#### What does the distribution look like over the years?

**Boxplot view**

```{r BOXPLOT}

## boxplot
## need to set x axis as factor for boxplot
nba_sum$season_start <- as.factor(nba_sum$season_start) 

ggplot(nba_sum, aes(x=season_start, y=salary))+geom_boxplot()+
  scale_y_continuous(labels=comma)
```

Layer in average salary for each year, along with 1 std dev above and 1 std dev below average, so to see how interquartile range compares over time. (for normal distribution, would be stable relationship)

```{r BOXPLOT SD }
## set x axis from reference data to factor 
nba_avg_sal$season_start <- as.factor(nba_avg_sal$season_start) 

ggplot(nba_sum, aes(x=season_start, y=salary))+geom_boxplot()+
  geom_point(data=nba_avg_sal, aes(x=season_start, y=avg_sal), 
             shape=19, color='blue', size=2)+
  geom_point(data=nba_avg_sal, aes(x=season_start, y=avg_sd_one_low), 
             shape=19, color='red', size=2)+
  geom_point(data=nba_avg_sal, aes(x=season_start, y=avg_sd_one_high), 
             shape=19, color='green', size=2)+
  scale_y_continuous(labels=comma)
```

As the upper end of the salary range stretches over time, it pulls the average away from the median and widens the sd range so that it is not as useful a representation of spread in the data than IQR. At the same time, IQR is getting distorted by the upward pull of salaries. However, it accomodates this by adjusting the upper and lower bounds of the 50% in the middle, so that upper is farther above median than lower is below. Std dev is always balanced on either side of average. 

**Histogram view**

```{r HISTOGRAM, fig.height=12}

#ggplot(nba_sum, aes(x=salary))+geom_histogram()+
#  facet_wrap(.~season_start)

ggplot(nba_sum, aes(x=salary))+geom_histogram()+
  facet_grid(season_start~.)+
  scale_x_continuous(labels=comma)+theme_light()

```

Too many years to deal with. How about group into 5 yrs.

```{r FIVE YR GROUP}
## use starting yr of 5 yr period
nba_sum5 <- nba_avg_sal
nba_sum5$fiveyr <- NA

## set start row s plus 4 rows after f to get 5 rows
## set fiveyr var for lowest season_start yr in set
## increase s to s +5 to jump to next 5 rows
## repeat until end

s <- 1
f <- s+4
while(s<nrow(nba_sum5)){
  nba_sum5$fiveyr[c(s:f)] <- nba_sum5$season_start[s]
s <- s+5
f <- min(s+4,nrow(nba_sum5))
}
nba_sum5 <- nba_sum5 %>% select(season_start, fiveyr)
nba_sum5$season_start <- as.factor(nba_sum5$season_start)
## join

nba_sum5yr <- left_join(nba_sum, nba_sum5, by='season_start')

```

**Better view when grouped over 5 yrs. **

```{r HISTOGRAM 5YR, fig.height=10}
ggplot(nba_sum5yr, aes(x=salary))+geom_histogram()+
  facet_grid(fiveyr~.)+
  scale_x_continuous(labels=comma)+theme_light()
```

As shown with boxplot, the center shifts up a little over the years - maybe not that much different from inflation - but the tail gets stretched longer over time.

### Scottie Pippin

Scottie Pippin is considered to have established himself as one of the all-time greats while at Chicago Bulls in the 1990's. At the same time, reported to have relatively low salary.

#### Career

The data I' using has player names as 'player_id'. Luckily, his is easy to find.

```{r}
nba_pip <- nba_sal %>% filter(str_detect(player_id,'pipp'))
nba_pip[,c(2:4,7)]
```

```{r}
ggplot(nba_pip, aes(x=season_start, y=salary, fill=team))+geom_col()+
  geom_hline(yintercept = 1000000, linetype='dashed')+
  scale_y_continuous(labels=comma, expand=c(0,0))

```

(dotted line = $1M)

Looks like even in early years with the Bulls, his salary varied. Although pushing $1M and then jumping well above it.

#### Where does he rank in the league each year?

```{r RANK}
## rank players by salary for each yr
nba_sal_rank <- nba_sal %>% group_by(season_start) %>% mutate(
  sal_rank=rank(desc(salary))
)
## test rank for a given yr
nba_sal_rank_yr <- nba_sal_rank %>% filter(season_start==1990) %>%
  arrange(sal_rank)
```

```{r PIPPEN RANK}
nba_pip_rank <- nba_sal_rank %>% filter(str_detect(player_id,'pipp'))

ggplot(nba_pip_rank, aes(x=season_start, y=sal_rank, fill=team))+geom_col()+
  scale_y_continuous(expand=c(0,0))
```

Varies wildly from year to year, with early years getting progressively bad, then rectified for 1991-1993, but started slipping again, until late 1990s. Houston Rockets and especially Portland Trail Blazers were much better to Scottie than the Bulls!

**How does he compare to Jordan, say in 1990?**

```{r PIPPEN COMPARE TBL}
nba_rank_compareyr <- nba_sal_rank %>% filter(season_start==1990)
nba_rank_compare <- nba_rank_compareyr %>% filter(sal_rank==1 | str_detect(player_id,'pipp') | str_detect(player_id,'jord'))

nba_rank_compare[,c(2:4,7,8)]
```

```{r CHART PIPPEN COMPARE SAL}
ggplot(nba_rank_compare, aes(x=player_id, y=salary))+geom_col()+
  geom_hline(yintercept=median(nba_rank_compareyr$salary), linetype='dashed')+
  scale_y_continuous(labels=comma, expand=c(0,0))
```

(dotted line = 1990 average salary)

Pippen slightly above average but well below his comrade Jordan.

#### Where does he rank percentile-wise?

```{r PIPPEN BOX PLOT}
## filter for years Pippen played
nba_sal_pipyr <- nba_sal %>% filter(season_start>=1987 & season_start<=2004)

## convert season_start fields to factor for boxplot
nba_sal_pipyr$season_start <- as.factor(nba_sal_pipyr$season_start)
nba_pip$season_start <- as.factor(nba_pip$season_start)

ggplot(nba_sal_pipyr, aes(x=season_start, y=salary))+geom_boxplot()+
  #geom_hline(yintercept=nba_pip$salary)+ ## draws lines across chart
  geom_point(data=nba_pip, aes(x=season_start, y=salary), 
             shape=19, color='red', size=2)+
  scale_y_continuous(labels=comma)
```

#### Calculate actual percentile

Percentiles are just rank order converted to percent (300/400 = 75th percentile)

```{r PERCENTILE CALC}
## rank players by salary for each yr
nba_sal_rank <- nba_sal_rank %>% group_by(season_start) %>% mutate(
  sal_percent_rank=percent_rank(salary)
)

## test rank for a given yr
nba_sal_rank_yr <- nba_sal_rank %>% filter(season_start==1990) %>%
  arrange(sal_percent_rank)

```


**How does he compare to Jordan on Percentile Rank?**

```{r}
nba_rank_compareyr <- nba_sal_rank %>% filter(season_start==1990)
nba_rank_compare <- nba_rank_compareyr %>% filter(sal_rank==1 | str_detect(player_id,'pipp') | str_detect(player_id,'jord'))

nba_rank_compare[,c(2:4,7,8,9)]
```

** Percentile chart, all yrs**

```{r CHART PERCENTILE}
ggplot(nba_sal_rank, aes(x=sal_percent_rank, y=salary, color=season_start))+geom_point()+
  scale_y_continuous(labels=comma)
```

Can convert season start to factor in order to have individual colors, but this works ok considering the number of yrs involved.